Meeting the Challenges of Modeling Astrophysical Thermonuclear Explosions: \castro, \maestro, and the \amrex\ Astrophysics Suite

### 2017-11-01 astronum_2017/astronum-amrex.tex
final edits

\castro\ has been applied to core-collapse
supernovae~\cite{castro-ccsne}, radiative shock breakout in
supernovae~\cite{lovegrove:2017}, population III pair-instability
supernovae~\cite{castro-pairinstability}, the Chandra model for
SNe~Ia~\cite{ma:2013}, the sub-Chandra SNe Ia model~\cite{moll:2013}, and white dwarf mergers as a model for SNe
Ia~\cite{moll:2014,wdmergerI}. Recent work has focused on improving energy
conservation with gravitational and rotational sources, and new
integration methods that more accurately couple hydro and reactions.
For \maestro\ simulations that evolve from the subsonic regime to the
sonic regime, we have demonstrated the ability to restart the
calculations in \castro\ to continue the evolution into the sonic
regime~\cite{scidac-petascale,malone:2014} (in this case, for Chandra
model SNe Ia).

@@

\maestro\ and \castro\ share the same microphysics, available as the
\starkiller\ astrophysics GitHub
project\footnote{\url{https://github.com/StarKiller-astro/microphysics/}}.
This includes equations of state and nuclear reaction
networks\footnote{Several of the reaction network righthand sides and
  the EOS were provided from Frank Timmes' {\em cococubed} software
  instruments page
  \url{http://cococubed.asu.edu/code_pages/codes.shtml}.  We thank him
  for making them available.}.  The reaction networks are written such
that the rates and integration strategy are decoupled, allowing us to
change the integration strategy for a given set of rates.  They are
also written to be threadsafe and with GPUs in mind (more on that
below).  The goal of \starkiller\ is to create a set of community
microphysics solvers that can be used in a variety of nuclear
astrophysics codes, not just those discussed here.

@@

Figure~\ref{fig:castro-scaling} shows \castro\ scaling behavior for a
pure hydro problem with a real EOS and a hydro + self-gravity problem
with a real EOS.  All tests were run on OLCF Titan.  The AMD
processors in Titan can be used in an 8 or 16 cores / node
configuration.  For all these runs, we ran in the 16 cores / node
mode.  We see excellent
scaling behavior for the pure hydro problem.  For the self-gravity
problem, we scale very well with a uniform grid, but with a factor of 4
refinement for the AMR level on top of this, we observe degraded
performance at higher core counts.  This is a challenging problem, as
only 3\% of the domain volume is refined.  For a given core count,
there are a number of different combinations of MPI ranks and OpenMP
threads we can use.  In general, with multilevel problems, we found
the best performance with fewer MPI ranks and more OpenMP threads.

@@

Figure~\ref{fig:knl-gpu-scaling} shows the performance of
\starlord\ on the Summitdev platform at OLCF\footnote{Summitdev
  consists of 2 IBM Power8 processors and 4 NVIDIA Pascal GPUs per
  node.}.  The highest GPU count (128) corresponds to 32 nodes on
Summitdev.  We see a nearly linear speedup with the number of GPUs,
indicating good weak scaling on this machine. A single P100 GPU achieves
a performance approximately 2.5 times that of the 20 Power8 cores. Efforts are underway to
complete the port of the GPU developments into \castro.  This figure
also shows our performance on the Intel Xeon Phi manycore processors (using
NERSC Cori).  This is for the same problem as the self-gravity test on
Titan.  The Intel Xeon Phi chips in Cori
have 68 cores than can be run with 1, 2, or 4 hardware threads each.
For all these runs, we ran with 4 threads / core (272
threads / chip).

\begin{figure}[t]
\centering
\includegraphics[width=0.48\linewidth]{CUDA-Fortran-VODE-Integrator}
\begin{minipage}[b]{0.48\linewidth}
\caption{\label{fig:cudaode} GPU speed up over single CPU core for two
  reaction networks.  The 13-isotope alpha-chain is a
  potential network for the sub-Chandra simulations.
  The 9-isotope Urca network is used for our Urca simulations.  A
  variety of grid sizes were used, $16^3$, $32^3$, $64^3$, and
  $128^3$.  In all cases, we see significant speed up on the GPU
  vs.\ a single core.  These speed-up numbers are from our testbed
  workstation with a Titan X Pascal board and Intel Xeon X5650 processors.}
\end{minipage}
\end{figure}

A parallel effort is porting our microphysics---in particular
reaction networks---to GPUs.  Our strategy is to do the entire ODE
integration on the GPU, i.e., the data for a patch of zones is passed
to the GPU, all righthand size and Jacobian evaluations and the
timestepping itself is done on the GPU, and once the burning in all
zones is completed, we access the data as needed on the CPU.  To
enable this, we ported our workhorse ODE integrator (VODE~\cite{vode})
to CUDA Fortran. This required extensive rewrites of the internals of
VODE, which was originally written using Fortran 77 syntax unsupported
by CUDA Fortran. Accelerating VODE with CUDA Fortran has proven
successful, and we now see significant performance gains with the CUDA
version of our reaction networks, even for moderate-sized networks (a
13-isotope standard network).  Figure~\ref{fig:cudaode} shows the
speed-ups on a GPU vs.\ single CPU core.

>>>>

\castro\ has been applied to core-collapse
supernovae~\cite{castro-ccsne}, radiative shock breakout in
supernovae~\cite{lovegrove:2017}, population III pair-instability
supernovae~\cite{castro-pairinstability}, the Chandra model for
SNe~Ia~\cite{ma:2013}, the sub-Chandra SNe Ia model~\cite{moll:2013},
and white dwarf mergers as a model for SNe
Ia~\cite{moll:2014,wdmergerI}.  For \maestro\ simulations that evolve
from the subsonic regime to the sonic regime, we have demonstrated the
ability to restart the calculations in \castro\ to continue the
evolution into the sonic regime~\cite{scidac-petascale,malone:2014}
(in this case, for Chandra model SNe Ia).

@@

\maestro\ and \castro\ share the same microphysics, available as the
\starkiller\ Microphysics GitHub
project\footnote{\url{https://github.com/StarKiller-astro/microphysics/}}.
This includes equations of state and nuclear reaction
networks\footnote{Several of the reaction network righthand sides and
  the EOS were provided from Frank Timmes' {\em cococubed} software
  instruments page
  \url{http://cococubed.asu.edu/code_pages/codes.shtml}.  We thank him
  for making them available.}.  The reaction networks are written such
that the rates and integration strategy are decoupled, allowing us to
change the integration strategy for a given set of rates.  They are
also written to be threadsafe and with GPUs in mind (more on that
below).  The goal of \starkiller\ is to create a set of community
microphysics solvers that can be used in a variety of nuclear
astrophysics codes, not just those discussed here.

@@

Figure~\ref{fig:castro-scaling} shows \castro\ scaling behavior for a
pure hydro problem with a real EOS and a hydro + self-gravity problem
with a real EOS.  All tests were run on OLCF Titan.  The AMD
processors in Titan can be used in an 8 or 16 cores / node
configuration.  For all these runs, we ran in the 16 cores / node
mode.  We see excellent scaling behavior for the pure hydro problem to
$\mathcal{O}(10^5)$ cores.  For the self-gravity problem, we scale
very well with a uniform grid, but with a factor of 4 refinement for
the AMR level on top of this, we observe degraded performance at
higher core counts.  This is a challenging problem, as only 3\% of the
domain volume is refined.  For a given core count, there are a number
of different combinations of MPI ranks and OpenMP threads we can use.
In general, with multilevel problems, we found the best performance
with fewer MPI ranks and more OpenMP threads.

@@

Figure~\ref{fig:knl-gpu-scaling} shows the performance of
\starlord\ on the Summitdev platform at OLCF\footnote{Summitdev
  consists of 2 IBM Power8 processors and 4 NVIDIA Pascal GPUs per
  node.}.  The highest GPU count (128) corresponds to 32 nodes on
Summitdev.  We see a nearly linear speedup with the number of GPUs,
indicating good weak scaling on this machine. A single P100 GPU achieves
a performance approximately 2.5 times that of the 20 Power8 cores. Efforts are underway to
complete the port of the GPU developments into \castro.  This figure
also shows our performance on the Intel Xeon Phi manycore processors (using
NERSC Cori).  This is for the same problem as the self-gravity test on
Titan.  The Intel Xeon Phi chips in Cori
have 68 cores than can be run with 1, 2, or 4 hardware threads each.
For all these runs, we ran with 4 threads / core (272
threads / chip).  We have not yet focused on optimizing \castro\ for the
Intel Xeon Phi architecture.

\begin{figure}[t]
\centering
\includegraphics[width=0.48\linewidth]{cuda-vode-speedup}
\begin{minipage}[b]{0.48\linewidth}
\caption{\label{fig:cudaode} GPU speed up over single CPU core for three
  different reaction networks.
  reaction networks.  The 13-isotope alpha-chain is a
  potential network for the sub-Chandra simulations;
  the 9-isotope Urca network is used for our Urca simulations;
  and the 3 isotope carbon network was used for our original \maestro\
  white dwarf convection calculations.  A
  variety of grid sizes were used, $16^3$, $32^3$, $64^3$, and
  $128^3$.  In all cases, we see a good  speed up on the GPU
  vs.\ a single core.  These speed-up numbers are from the
  Summitdev machine, comparing a single Tesla P100 GPU to a single Power8 CPU core.}
\end{minipage}
\end{figure}

A parallel effort is porting our microphysics---in particular
reaction networks---to GPUs.  Our strategy is to do the entire ODE
integration on the GPU, i.e., the data for a patch of zones is passed
to the GPU, all righthand size and Jacobian evaluations and the
timestepping itself is done on the GPU, and once the burning in all
zones is completed, we access the data as needed on the CPU.  To
enable this, we ported our workhorse ODE integrator (VODE~\cite{vode})
to CUDA Fortran. This required extensive rewrites of the internals of
VODE, which was originally written using Fortran 77 syntax unsupported
by CUDA Fortran. Accelerating VODE with CUDA Fortran has proven
successful, and we now see significant performance gains with the CUDA
version of our reaction networks, even for moderate-sized networks (a
13-isotope standard network).  Figure~\ref{fig:cudaode} shows the
speed-ups on a GPU vs.\ single CPU core.  The main issue with scaling
is running out of local stack memory per thread with larger networks.
Reducing the memory footprint is a near-term goal for this work.  This
test problem also shows a lot of thread divergence due to the widely
differing thermodynamic conditions in the zones that are burning.
